\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-11-23 paper.tex
incorporated comments of review meeting

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords, encryption
keys). IoT devices typically have limited batteries, processing power and
memory. These properties make it hard to add any additional service, including
security.

@@

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, through the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service or even access the network in a legitimate way. For each of these
attacks an algorithm needs to look for patterns or anomalies. An IDS for IoT
devices therefore requires a large number of implemented algorithms for an
adequate coverage.

@@

Current developments in ID focus on programming frameworks that structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. A key
shortcoming of prior ID frameworks, is that they don't offer a way to optimize
the usage of resources. A solution that supports the integration of multiple
algorithms on IoT devices needs to avoid accumulating their impact on the
available resources.

A second omission is support for systematic reuse to create different
configurations heterogeneous devices. Addressing this in a transparent and
automated way offers the potential to significantly reduce the effort of
developing an IDS.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: anomaly detection,
signature or misuse detection, and specification-based detection. In
\cite{alrajeh2013intrusion} the authors agree with this typology and add the
notion of hybrid intrusion detection systems and cross layer intrusion
detection systems as recent advances in research. Although these additional
categories seem promising, the authors admit that the impact on
resource-constrained devices will be too much. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Specification-based detection] uses constraints that describe the
  correct behavior of the system. Although that this approach can detect
  previously unknown attacks, it can sometimes not detect specific attacks that
  anomaly or misuse detection can.

@@

Algorithms that fit this classification are typically platform independent.
They analyze network messages, looking for patterns that can reveal intrusions
and all deal with the same information: other devices in the network. To be
able to formally describe ID algorithms and target the heterogeneous IoT, \NAME
requires ID algorithms to expose these properties.

@@

  \item[Reimplementing] the algorithms manually to eliminate these issues is
  not a one-time cost. For example due to the release of a new version of an
  algorithm, or simply a reselection of algorithms due to a changing risk
  analysis. Also, changing, or simply implementing the algorithms from scratch,
  holds the risk of making mistakes.

@@

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms.
But at the same time they conflict with their target environment, the IoT\@.
\NAME provides a framework to efficiently implement an IDS for the IoT and
offers a solution for the identified problems. It consists of a DSL, source
code generator and software library. Figure \ref{fig:design} visualises the
overall design.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=7mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)  [resource]                                          {DSL};
  \node (cg)   [representation,right=of dsl,align=center]          {Code\\Generator};
  \node (code) [resource,right=of cg]                              {Code};
  \node (lib)  [resource,below=of code,align=center,yshift=0.5cm]  {Software\\Library};

  \draw [->] (dsl) -> (cg)   node [midway,above=25pt]              {intent};
  \draw [->] (cg)  -> (code) node [midway,above=15pt,align=center] {platform\\ + language};

  \draw [rounded corners,->]
        ($ (lib.west) $)
        -- ++(-0.5cm, 0cm)
        -- ++( 0cm,   1cm)
        -- ($ (code.west) + (0cm,-0.22cm) $);

\end{tikzpicture}
}
\caption{Overall Design}
\label{fig:design}
\end{figure}

@@

A DSL and code generator defines a development strategy. This implies choices
with respect to flexibility, update cost and development effort. We will first
position \NAME and motivate its design with respect to these trade offs.
Second, we will look at the principal design components of the DSL and how it
deals with loops and common data. Third, we discuss the structure of the code
generator and introduce its software library, that offers a way to deal with
the network access issue.

@@

\newcommand{\ball}[5][above,yshift=+5pt]{
\tdplottransformmainscreen{#2}{#3}{#4}
\shadedraw[tdplot_screen_coords, ball color=lightgray,draw=lightgray,shading angle=-90]
  (\tdplotresx,\tdplotresy) circle (0.20)
  node[text=gray,font=\footnotesize,#1]{#5};
}

\begin{figure}[hb]
  \centering
  \scalebox{.85}{
\tdplotsetmaincoords{110}{-10}
\begin{tikzpicture}[scale=1,tdplot_main_coords]
  % axis
  \draw[thick,->] (0,0,0) -- (7,0,0) node[anchor=north east,xshift=10mm]{$flexibility$};
  \draw[thick,->] (0,0,0) -- (0,7,0) node[anchor=north west]{$update\ cost$};
  \draw[thick,->] (0,0,0) -- (0,0,4) node[anchor=south]{$development$};
  % points
  \coordinate (O)        at (0,0,0);
  \coordinate (ideal)    at (6,0,0);
  \coordinate (vm)       at (2,2,2);
  \coordinate (modular)  at (4,4,4);
  \coordinate (reconfig) at (4,2,4);
  \coordinate (full)     at (6,6,4);
  \coordinate (cg)       at (6,4,2.5);
  % "lines"
  \draw[dashed] (2,0,0)    -- (2,2,0) -- (0,2,0);
  \draw[dashed] (4,0,0)    -- (4,4,0) -- (0,4,0);
  \draw[dashed] (6,0,0)    -- (6,6,0) -- (0,6,0);
  \draw[dashed] (vm)       -- (2,2,0);
  \draw[dashed] (modular)  -- (4,4,0);
  \draw[dashed] (full)     -- (6,6,0);
  \draw[dashed] (reconfig) -- (4,2,0);
  \draw[dashed] (cg)       -- (6,4,0);
  % "balls"
  \ball{6}{0}{0}{ideal}
  \ball{2}{2}{2}{virtual machine}
  \ball[left,xshift=-5pt]{4}{4}{4}{\shortstack{modular\\native image}}
  \ball{4}{2}{4}{\shortstack{reconfigurable\\virtual machine}}
  \ball[right,xshift=+5pt,yshift=+5pt]{6}{6}{4}{full native image}
  %\ball[right,xshift=+5pt,text=black]{6}{4.5}{3}{code generation};
  % 5. cg
  \tdplottransformmainscreen{6}{4}{2.5}
  \shadedraw[tdplot_screen_coords, ball color=gray,draw=black,shading angle=-90]
    (\tdplotresx,\tdplotresy) circle (0.20)
    node[right,xshift=+5pt,text=black,align=center,font=\footnotesize]{code generation\\+ functional code fusion};
  % "box around our solution" ;-)
  \tdplottransformmainscreen{0}{0}{0}
  \draw[tdplot_screen_coords] (4.95,1) rectangle (8.5,0.25);

\end{tikzpicture}
}
\caption{Classification of Development Strategies for Wireless Devices (based on \cite{sadilek2007energy})}
\label{fig:positioning}
\end{figure}

Figure \ref{fig:positioning} shows a classification of possible development
strategies for projects involving wireless devices. Three properties can
classify development strategies \cite{sadilek2007energy}: flexibility, update
cost and development effort.

  \item[Flexibility] is a measure for the number of applications one can deploy
  on a device without physical access.

  \item[Update cost] is equivalent to the energy required to deploy a new
  version of an application on a device.

  \item[Development effort] covers the time to write, test and deploy an
  application.

The \emph{ideal} situation is that were one ends up with a maximal flexibility
and no update nor development cost. \NAME starts from the full native image,
allowing full flexibility, and aims to reduce both the update and development
cost: first, by applying functional code fusion, \NAME optimizes the code
organization, which has a positive effect on the code size and hence also on
the update cost. Second, by enabling reuse on a functional level and automating
the writing phase, the development cost is also lowered.

@@

@@

All algorithms deal with information about other devices in the network. To
centralize this information \NAME provides a unified \ttt{nodes} concept as
part of its DSL and exposes it as an object to the implementer.

Algorithms can extend nodes with their own properties. This way, only one
instance for every known device will be present in memory. Finally, nodes can
define a scope for a function to execute in. This way, nodes also offer a way
to i.e.\ iterate all known devices, again without the need for a loop.

@@

@@

\end{tikzpicture}
}
\caption{Overview of \NAME's code generation process}
\label{fig:code-generation}
\end{figure}

@@

  \draw [composite]  (nodes)    -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}
\caption{Partial Semantic Model for \NAME. Shaded classes indicate the \emph{event/scheduling} construction around functions to allow interpretation.}
\label{fig:meta-model}
\end{figure}

@@

@@

% table goes here, because a spanning table is placed on the next page, relative
% to where it's defined.
\begin{table*}[b]
  \centering
  \begin{tabular}{llrrrrrrr}
\toprule
              &            & base  & \multicolumn{2}{c}{watchdog} & \multicolumn{2}{c}{reputation} & \multicolumn{2}{c}{both} \\
\midrule
image
size (bytes)  & manual     & 10466 & 15530 &  +48\% & 13306 &  +27\% & 18334 &  +75\%\\
              & generated  & 10466 & 18352 &  +75\% & 16376 &  +56\% & 20998 & +101\%\\
\cmidrule{2-9}
              & difference &     0 &  2822 &  +18\% &  3070 &  +23\% &  2664 &  +15\%\\
\midrule
\multirow{2}{*}{\pbox{\textwidth}{total packets sent\\in a window of 90s}}
              & manual     &    20 &    51 & +155\% &    32 &  +60\% &    63 & +215\%\\
              & generated  &    20 &    49 & +145\% &    32 &  +60\% &    55 & +175\%\\
\cmidrule{2-9}
              & difference &     0 &    -2 &   -4\% &     0 &   0\% &    -8 &   -13\%\\
\midrule
\multirow{2}{*}{\pbox{\textwidth}{total bytes sent\\in a window of 90s}}
              & manual     &   476 &  1933 & +306\% &   860 &  +81\% &  2317 & +387\%\\
              & generated  &   476 &  1897 & +299\% &   884 &  +86\% &  2161 & +354\%\\
\cmidrule{2-9}
              & difference &     0 &   -36 &   -2\% &    24 &   +3\% &  -156 &   -7\%\\
\midrule
\multirow{2}{*}{\pbox{\textwidth}{event loop\\average duration ($\mu$s)}}
              & manual     &    48 &    94 &  +96\% &    88 &  +83\% &   149 & +210\%\\
              & generated  &    48 &   121 & +152\% &   121 & +152\% &   138 & +188\%\\
\cmidrule{2-9}
              & difference &     0 &    27 &  +29\% &    33 &  +38\% &   -11 &   -7\%\\
\bottomrule
  \end{tabular}
  \caption{Results and Side-by-Side Comparison.}
  \label{tbl:results}
\end{table*}

This section discusses a prototype implementation and evaluates its
performance. We first present the architecture of this implementation and
consider the impact it has from a development point of view. Next we use this
implementation to quantify its merits using a side-by-side comparison of
generated code versus a manual implementation. We consider experimental results
regarding the image size, usage of the network and execution time of the
event-loop. We also take a more analytical approach and look at the handling of
common data and the size of the source code.

\subsection{Architecture}

Figure \ref{fig:architecture} shows the architecture of the implementation.

\begin{figure}[ht]
  \centering
  \scalebox{.85}{
\begin{tikzpicture}[
  node distance=10mm and 20mm,
  ,>=stealth,thick,black!50,text=black,
  every new ->/.style={shorten >=1pt},
]
  \node (dsl)      [user,align=center]                                     {DSL};
  \node (code)     [generated,right=of dsl]                                {code};
  \node (image)    [generated,right=of code]                               {image};

  \node (nodes)    [dom,above right=of dsl,xshift=-20mm]                   {Nodes};
  \node (platform) [platform,below right=of dsl,align=center,xshift=-20mm] {Platform \&\\Language};
  \node (lib)      [native,below right=of code,align=center,xshift=-20mm]  {Software\\Library};

  \draw [->] (dsl)  -> (code)  node [midway,above=2pt] {generation};
  \draw [->] (code) -> (image) node [midway,above=2pt] {compilation};

  \draw [->,dashed] (nodes.south)    -> ++(0,-15mm);
  \draw [->,dashed] (platform.north) -> ++(0,15mm);
  \draw [->,dashed] (lib.north)      -> ++(0,15mm);
\end{tikzpicture}
}
  \caption{Architecture of Prototype Implementation}
  \label{fig:architecture}
\end{figure}

It illustrates the generation of the DSL into compilable code, using the Nodes
domain information, a platform and language description and the native software
library. The applied grayscale indicates the components' platform dependency:
the DSL is platform and language independent and deals with subsystems, such as
the network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
The platform and language descriptions define the target platform and language
and convert the abstract constructions in actually supported code. The software
library is standard native code, targeting the platform and language. Finally,
the code and image are artifacts of the automated generation and compilation
process. They aren't a product of manual work and therefore aren't directly
dependent on the platform.

@@

To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
micro-controller and the Digi XBee S2 ZigBee module. A basic library wrapped
the technical calls to the hardware components in more easy to use functions.

@@

We evaluated three criteria: the image size of the resulting compiled code, the
network usage in number of packets and bytes and the time required to perform
once cycle of the event loop. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

@@

Table \ref{tbl:results} also compares the two situations by subtracting the
manual case from the generated case, showing the impact of the code generation.
A relative value for this difference, to the manual case, is also included.

@@

Secondly, we see that the software library that comes with the generated code
adds to the size of the resulting image. But the cost is close to constant and
is lower for both algorithms together. With roughly 3KB of additional overhead,
or 15\% in this simple case, the impact is small. Taking into account the
available memory on the ATMega1284p, this is only 2.3\% of the available
resources.

Thirdly, when turning to the goal to reduce the network usage, we see that both
the number of packets as well as the amount of bytes sent have dropped by 13\%
and 7\% respectively.

Finally, focusing on the execution time, both algorithms by themselves add an
equal amount of time to the processing of a single event loop cycle. But when
combined, the event loop is about 7\% faster compared to the manual case.

@@

Looking at the generated source code is another way to validate the framework.
A simple metric is comparing the amount of code versus the size of the formal
description of IDS algorithms. This can be an indication of the development
effort and complexity. In this experiment, implementing both algorithms
manually results in 637 lines of C code. The corresponding \NAME implementation
results in 156 lines of DSL code.

@@

@@

Applying code generation to the context of ID, often focusses on the generation
of native code to avoid the interpretation of descriptions of patterns to look
for. I2S \cite{charitakis2003code} is an example that generates code for SNORT
\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
targets a specific platform.

@@

Adequate intrusion detection for IoT devices requires multiple algorithms to
operate at the same time. Recent literature describes different algorithms, but
few are readily available for direct application when implementing an IDS for
IoT devices. Reselection of algorithms, due to changing requirements or
security threats, requires a flexible way to incorporate the different
algorithms. This eliminates manual implementation and optimization.

Wireless resource-constraint devices, such as IoT devices, require that
software respects their restrictions. Adding multiple intrusion detection
algorithms requires fusion of these algorithms on a functional level, to reduce
execution inefficiencies, redundant memory usage and superfluous network access.

\NAME is a framework consisting of a DSL, a source code generator and
accompanying software library. Using the DSL, intrusion detection algorithms
can be formally described in terms of functionality. During source code
generation, functional code fusion interprets the functional meaning of the
algorithms and merges common functionality and shared data. The software
library hosts common functionality, such as network message parsing, sent
message aggregation and event scheduling. This allows the source code generator
to focus on the optimization of the organization of the functionality.

Initial experiments confirm the theoretical positive impact on resource usage.
The platform independent DSL enables generation of source code targeting
different platforms and languages, which allows high degrees of reuse in
heterogenous environments and different configurations.

Future research will focus on the integration of applications with the
generated intrusion detection systems. We see opportunities for applications
who detect intrusion, to actively and dynamically respond and avoid malicious
payloads.

>>>>

The internet of things (IoT) holds great potential to positively influence our
daily work and life through domotics, assisted living, e-health and enhanced
learning \cite{atzori2010internet}. But with this potential the IoT also
presents a significant threat: by opening our smart homes and personal data to
all of these interconnected devices, we also open them to everyone who is able
to break the virtual locks that protect them (i.e.\ passwords, encryption
keys). Implementation of security measures by itself is hindered by the fact
that IoT devices typically have limited batteries, processing power and memory.

@@

Attackers have access to a large set of possible attack vectors
\cite{aschenbruck2012security}, ranging from the wireless network, through the
medium access control and routing, up to the device itself. Each level presents
different opportunities to manipulate data, eavesdrop, perform a form of denial
of service. For each of these attacks an algorithm needs to look for patterns
or anomalies. An IDS for IoT devices therefore requires a large number of
implemented algorithms for an adequate coverage.

@@

Current developments in ID focus on programming frameworks that structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. A key
shortcoming of prior ID frameworks, is that they don't offer a way to optimize
the usage of resources. A solution that supports the integration of multiple
algorithms on IoT devices needs to avoid accumulating their impact on the
available resources. A second omission is support for systematic reuse to
create different configurations heterogeneous devices. Addressing this in a
transparent and automated way offers the potential to significantly reduce the
effort of developing an IDS.

@@

There are different ways to construct a taxonomy for intrusion detection, but
common themes do appear. According to \cite{mishra2004intrusion} and
\cite{ioannis2007towards} there are three major categories: 1) anomaly
detection, 2) signature or misuse detection, and 3) specification-based
detection. In \cite{alrajeh2013intrusion} the authors agree with this typology
and add the notion of hybrid intrusion detection systems and cross layer
intrusion detection systems as recent advances in research. Although these
additional categories seem promising, the authors admit that the impact on
resource-constrained devices will be too much. We therefore focus on the three
categories agreed upon by most sources.

@@

  \item[Specification-based detection] uses constraints that describe the
  correct behavior of the system. Although this approach can detect previously
  unknown attacks, it can sometimes not detect specific attacks that anomaly or
  misuse detection can.

@@

Algorithms that fit this classification are typically platform independent.
They look for patterns in network messages that can reveal intrusions and, in
doing so, all deal with the same information: other devices in the network. To
be able to formally describe ID algorithms and target the heterogeneous IoT,
\NAME requires ID algorithms to expose these properties.

@@

  \item[Reimplementing] the algorithms manually to eliminate these issues is
  not a one-time cost. A release of a new version of an algorithm, or simply a
  reselection of algorithms, due to a changing risk analysis, can trigger it.
  Also, changing, or simply implementing the algorithms from scratch, holds the
  risk of making mistakes.

@@

\begin{figure*}[b]
  \centering
  \scalebox{1}{
\begin{tikzpicture}[
  node distance=9mm,
  ,>=stealth,very thick,black!50,text=black,
  every new ->/.style={shorten >=1pt}
]
  \node (dsl)   [user]                                      {DSL};
  \node (cg)    [representation,right=of dsl,align=center]  {Code\\Generator};
  \node (code)  [generated,right=of cg]                     {code};
  \node (comp)  [representation,right=of code]              {Compiler};
  \node (image) [generated,right=of comp]                   {image};

  \node (nodes) [dom,above=of cg, align=center]             {Nodes\\Domain};
  \node (plat)  [platform,below=of cg,align=center]         {Platform \&\\Language};          
  \node (lib)   [native,below=of comp,align=center]         {Software\\Library};

  \draw [->] (dsl)   -> (cg);
  \draw [->] (nodes) -> (cg);
  \draw [->] (plat)  -> (cg);
  \draw [->] (cg)    -> (code);
  \draw [->] (code)  -> (comp);
  \draw [->] (lib)   -> (comp);
  \draw [->] (comp)  -> (image);

\end{tikzpicture}
}

\caption{The high level design of \NAME, showing the resources, processes and
artifacts. A grayscale background represents the level of platform dependency,
with darker shades indicating a higher dependency. Both the source code and the
installable image are artifacts.}

\label{fig:design}
\end{figure*}

@@

The previous sections identified loops, common data and the network as three
important code-level problems. These problems are inherent to ID algorithms.
But at the same time they conflict with their target environment, the IoT\@.
\NAME provides a framework to efficiently implement an IDS for the IoT and
offers a solution for the identified problems. It consists of a 1) DSL, 2)
source code generator and 3) software library. Figure \ref{fig:design}
visualizes the overall design.

% figure design is moved up to end up on this page

@@

A DSL and code generator defines a development strategy. This has implications,
both on what is possible and how it can be achieved. We will first position
\NAME and motivate its design with respect to these trade offs. Second, we will
look at the principal design components of the DSL. Third, we discuss the
structure of the code generator and introduce its software library, that offers
a way to deal with for example the network access issue.

@@

The introduction of a code generator in the development process changes the
traditional way of working. Most importantly because the actual source code is
no longer the driving force that can be controlled by developers. It is
literally an intermediate representation between code generator and compiler.

A code generator constructs this representation using three components, as
shown in figure \ref{fig:design}: 1) the description of the algorithm in the
DSL, 2) information about the domain, in this case the Nodes domain and 3)
information about the platform and target language. At the level of the
compiler, a software library is introduced to offload specialized
implementations and allow custom extensions.

The applied grayscale indicates the components' platform dependency: the DSL is
platform and language independent and deals with subsystems, such as the
network, in an abstract way. The Nodes domain implements some of these
abstracted subsystems, but still in a platform and language independent way.
The platform and language descriptions define the target platform and language
and convert the abstract constructions in actually supported code. The software
library is standard native code, targeting the platform and language. Finally,
the code and image are artifacts of the automated generation and compilation
process. They aren't a product of manual work and therefore aren't directly
dependent on the platform.

@@

Besides avoiding these problems, \NAME's DSL also provides a higher level of
abstraction with respect to memory management and data access.

This section therefore discusses \NAME's approach to 1) loops, 2) nodes domain,
3) unified messaging, 4) language support and illustrates these with a complete
example.

@@

Most, if not all, ID algorithms deal with information about other devices in
the network. To centralize this information \NAME provides a unified
\ttt{nodes} concept as part of its DSL and exposes it as an object to the
implementer.

Algorithms can extend nodes with their own properties. This way, only one
instance for every known device will be present in memory. Finally, nodes can
define a scope for a function to execute in. This way, nodes also offer a way
to i.e.\ iterate all known devices, again without the need for an explicit loop.

@@

\subsubsection{Language Support}
\label{language-support}

Using a platform independent language also offers the possibility to relief the
implementer of explicit memory management and low-level data structures. From a
language perspective, \NAME offers several programming constructs, typically
found in modern programming languages with higher levels of abstraction:
optional typing, object orientation, pattern matching, aspect orientation.

@@

\end{tikzpicture}
}
\caption{\NAME's code generation process}
\label{fig:code-generation}
\end{figure}

@@

  \draw [composite]  (nodes)    -- (scope);
  \draw [dependency] (strategy) -- (function);
\end{tikzpicture}
}

\caption{Partial Semantic Model for \NAME. Shaded classes indicate the
\emph{event/scheduling} construction around functions to allow interpretation
of the intended functionality.}

\label{fig:meta-model}
\end{figure}

@@

% table goes here, because a spanning table is placed on the next page, relative
% to where it's defined.
\begin{table*}[b]
  \centering
  \begin{tabular}{llrrrrrrr}
\toprule
              &            & base  & \multicolumn{2}{c}{watchdog} & \multicolumn{2}{c}{reputation} & \multicolumn{2}{c}{both} \\
\midrule
\multirow{2}{*}{\pbox{\textwidth}{total packets sent\\in a window of 90s}}
              & manual     &    20 &    51 & +155\% &    32 &  +60\% &    63 & +215\%\\
              & generated  &    20 &    49 & +145\% &    32 &  +60\% &    55 & +175\%\\
\cmidrule{2-9}
              & difference &     0 &    -2 &   -4\% &     0 &   0\% &    -8 &   -13\%\\
\midrule
\multirow{2}{*}{\pbox{\textwidth}{total bytes sent\\in a window of 90s}}
              & manual     &   476 &  1933 & +306\% &   860 &  +81\% &  2317 & +387\%\\
              & generated  &   476 &  1897 & +299\% &   884 &  +86\% &  2161 & +354\%\\
\cmidrule{2-9}
              & difference &     0 &   -36 &   -2\% &    24 &   +3\% &  -156 &   -7\%\\
\midrule
\multirow{2}{*}{\pbox{\textwidth}{event loop\\average duration ($\mu$s)}}
              & manual     &    48 &    94 &  +96\% &    88 &  +83\% &   149 & +210\%\\
              & generated  &    48 &   121 & +152\% &   121 & +152\% &   138 & +188\%\\
\cmidrule{2-9}
              & difference &     0 &    27 &  +29\% &    33 &  +38\% &   -11 &   -7\%\\
\midrule
image
size (bytes)  & manual     & 10466 & 15530 &  +48\% & 13306 &  +27\% & 18334 &  +75\%\\
              & generated  & 10466 & 18352 &  +75\% & 16376 &  +56\% & 20998 & +101\%\\
\cmidrule{2-9}
              & difference &     0 &  2822 &  +18\% &  3070 &  +23\% &  2664 &  +15\%\\
\bottomrule
  \end{tabular}
  \caption{Results and Side-by-Side Comparison.}
  \label{tbl:results}
\end{table*}

@@

In this section we present a prototype implementation and evaluates its
performance. We first present the implementation consisting of the hardware
platform, network and software stack. Next we use this implementation to
quantify its merits using a side-by-side comparison of generated code versus a
manual implementation. We consider experimental results regarding the image
size, usage of the network and execution time of the event-loop. We also take a
more analytical approach and look at the handling of common data and the size
of the source code.

@@

To evaluate the prototype we used a system based on the Atmel ATMEGA1284p
micro-controller and the Digi XBee S2 ZigBee module. A basic software library
wraps the technical calls to the hardware components in more easy to use
functions.

@@

We evaluated three criteria: the network usage in number of packets and bytes,
the time required to perform once cycle of the event loop and the image size of
the resulting compiled code. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

@@

Table \ref{tbl:results} also compares the two situations by subtracting the
manual case from the generated case, showing the impact of the code generation.
A relative value for this difference, to the manual case, is also included.
Four observations allow us to evaluate the overall performance of \NAME's
approach.

@@

Secondly, when turning to the goal to reduce the network usage, we see that both
the number of packets as well as the amount of bytes sent have dropped by 13\%
and 7\% respectively.

Thirdly, focusing on the execution time, both algorithms by themselves add an
equal amount of time to the processing of a single event loop cycle. But when
combined, the event loop is about 7\% faster compared to the manual case.

Finally, we see that the software library that comes with the generated code
adds to the size of the resulting image. But the cost is close to constant and
is lower for both algorithms together. With roughly 3KB of additional overhead,
or 15\% in this simple case, the impact is small. Taking into account the
available memory on the ATMega1284p, this is only 2.3\% of the available
resources.

@@

Looking at the generated source code is another way to validate the framework.
A simple metric is comparing the amount of code versus the size of the formal
description of IDS algorithms. This can be an indication of the development
effort and complexity. In this experiment, implementing both algorithms
manually results in 637 lines of C code. The corresponding \NAME implementation
results in 156 lines of DSL code. This is largely thanks to the higher level
coding constructs, as introduced in section \ref{language-support} dealing with
\NAME's language support.

@@

Although not a real DSL, nesC \cite{gay2003nesc} is a general purpose language
that shares some typical concepts with \NAME, such as the high degree of
analysis during code generation to optimize execution. The biggest difference
is that \NAME starts from the intended functionality and can optimize certain
constructions that can't be interpreted from more technical source code.

@@

Applying code generation to ID, often focusses on the generation of native code
to avoid the interpretation of descriptions of patterns to look for. I2S
\cite{charitakis2003code} is an example that generates code for SNORT
\cite{roesch1999snort} rules. To optimize this highly efficient code, it often
targets a specific platform.

@@

Adequate intrusion detection for IoT devices requires multiple algorithms to
operate at the same time. Frequent reselection, e.g. due to changing
requirements or security threats, requires a flexible way to incorporate the
different algorithms.

Adding multiple intrusion detection algorithms to wireless resource-constraint
devices requires fusion of these algorithms on a functional level. This can
reduce execution inefficiencies, redundant memory usage and superfluous network
access.

\NAME is a framework consisting of a DSL, a source code generator and software
library. Using the DSL, intrusion detection algorithms can be formally
described. The code generator applies functional code fusion to interpret the
intent of algorithms and merges common functionality and shared data. The
software library hosts common functionality, such as network message parsing
and event scheduling.

Initial experiments confirm the theoretical positive impact on resource usage:
grouping of messages lowers the number of packets by 13\%, while the total
number of bytes sent over the network is reduced by 7\%. Finally, the execution
time is also improved by 7\%.

The platform independent DSL enables generation of source code targeting
different platforms and languages, which allows high degrees of reuse with
heterogenous devices.

As a next step, we want to investigate the relation of \NAME's DSL to reactive
programming. Both clearly share common concepts with respect to
event-drivenness and the absence of e.g. explicit loops.

Future research will focus on the integration of applications with a generated
IDS. We see opportunities for applications who detect intrusion, to actively
and dynamically respond and avoid malicious payloads.

### 2014-11-23 tikz.tex
incorporated comments of review meeting

\tikzset{
  artifact/.style       = { box, font=\ttfamily },
  support/.style        = { box, font=\itshape },
  user/.style           = { artifact, fill=black!15!white },
  dom/.style            = { support,  fill=black!30!white },
  platform/.style       = { support,  fill=black!55!white, text=white },
  native/.style         = { support,  fill=black!70!white, text=white },
  generated/.style      = { artifact, border },
  resource/.style       = { artifact, border },
  representation/.style = { support, border, round }
}

>>>>

\tikzset{
  artifact/.style       = { box, font=\ttfamily },
  support/.style        = { box, font=\itshape },
  user/.style           = { artifact, fill=black!15!white },
  dom/.style            = { support,  fill=black!30!white },
  platform/.style       = { support,  fill=black!55!white, text=white },
  native/.style         = { support,  fill=black!70!white, text=white },
  generated/.style      = { artifact, border, dashed },
  resource/.style       = { artifact, border },
  representation/.style = { support, border, round }
}
